Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-01-05 chemotaxis/paper/draft1.tex
updated chemotaxis paper

Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
%
Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
%
Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
%
The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a constant vector which is not used as an embedding coordinate.

In the limit of infinite data ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplacian operator on the manifold $\mathcal{M}_d$ \cite{...}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approach the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
%
The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues we compute in diffusion maps by 
\begin{equation}
\mu_k = e^{-\tilde{\mu}_k t}.
\end{equation}

@@

\begin{itemize}
\item Discuss the issues: harmonics
\item Give simple example (Laplacian on strip)
\end{itemize}

@@

Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$.
%
The corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two largest (non-trivial) eigenvalues (the largest eigenvalue will always be $\mu_{0,0} = 1$ and correspond to a constant eigenfunction). 
%
In fact, if $L_x > 2 L_y$, then $\mu_{2, 0} > \mu_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are sorted by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other.

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
%
To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
\end{equation}
where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha} (i) , \hat{\beta}(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $w$ is a weighting function.
%
We will use Gaussian weights, 
%
\begin{equation}
w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right).
\end{equation}
%
We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
%
The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...}.

We then define the normalized leave-one-out cross-validation error as
\begin{equation} \label{eq:cv_error}
r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}(i) + \hat{\beta}(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
\end{equation}
%
%We will use a local linear function 
%\begin{equation}
%f_{k}(\phi_1, \dots, \phi_{k-1}) = a_0 + a_1 \phi_1 + \dots + a_{k-1} \phi_{k-1}
%\end{equation}
%where the coefficients $a_0, \dots, a_{k-1}$ are functions of $\phi_1, \dots, \phi_{k-1}$. 
%
%To find these coefficients, we want to minimize the local weighted sum of squares
%\begin{equation} \label{eq:opt_problem}
%\sum_{i=1}^n w_i(x) \left( \phi_{k}(x) - f_{k} (\phi_1(x), \dots, \phi_{k-1}(x)) \right)^2
%\end{equation}
%where $w$ is a weighting function.
%%
%We will use Gaussian weights, 
%%
%\begin{equation}
%w_i(x) = \exp \left( - \frac{\|\Phi_{k-1}(x) - \Phi_{k-1} (x_i) \|^2}{\epsilon^2} \right), \text{where } \Phi_{k-1}(x) = \begin{bmatrix}
%\phi_1(x) \\
%\vdots \\
%\phi_{k-1}(x)
%\end{bmatrix} .
%\end{equation}
%%
%We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(x)$.
%
%Let $\hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
%
%We then define the normalized leave-one-out cross-validation error as
%\begin{equation} \label{eq:cv_error}
%r_{k} = \sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i)) \right)^2
%\end{equation}
%\begin{equation} \label{eq:cv_error}
%r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\Phi_{k-1}(x_i)) \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (x_i) \right) }}
%\end{equation}
%
When $r_{k}$ is small, $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
%
A large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
%
Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we define $r_1 = 1$.

@@

\begin{figure}[t] 
\includegraphics[width=8cm]{detect_change_eigenvalues}
\caption{The first two eigenvalues from diffusion maps as a function of $\lambda$. Each point is averaged over $10$ simulation experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation. 
The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
\label{fig:detect_change}
\end{figure}

@@

To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
%
Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
%
We plot  
\begin{equation}
 \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
\end{equation}
when this ratio becomes small, it indicates that the data is effectively one-dimensional, as the second dimension is very small compared with the first. 
%
This is plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$.

>>>>

Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
%
Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
%
Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
%
The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.

In the limit of infinite data ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplacian operator on the manifold $\mathcal{M}_d$ \cite{...}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
%
The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues we compute in diffusion maps by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = e^{-\tilde{\mu}_k t}.
\end{equation}

@@

@@

Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
%
The eigenvalues of the Laplacian with Neumann boundary conditions are given by
\begin{equation}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
\end{equation}
where $x$ and $y$ denote the two coordinates of the rectangular domain. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold. 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction. 
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
%
To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
\end{equation}
where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $w$ is a weighting function.
%
We will use Gaussian weights, 
%
\begin{equation}
w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
\end{equation}
%
and take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
%
The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...}.

We then define the normalized leave-one-out cross-validation error for this local linear fit as
\begin{equation} \label{eq:cv_error}
r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}_k(i) + \hat{\beta}_k(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
\end{equation}
%
%We will use a local linear function 
%\begin{equation}
%f_{k}(\phi_1, \dots, \phi_{k-1}) = a_0 + a_1 \phi_1 + \dots + a_{k-1} \phi_{k-1}
%\end{equation}
%where the coefficients $a_0, \dots, a_{k-1}$ are functions of $\phi_1, \dots, \phi_{k-1}$. 
%
%To find these coefficients, we want to minimize the local weighted sum of squares
%\begin{equation} \label{eq:opt_problem}
%\sum_{i=1}^n w_i(x) \left( \phi_{k}(x) - f_{k} (\phi_1(x), \dots, \phi_{k-1}(x)) \right)^2
%\end{equation}
%where $w$ is a weighting function.
%%
%We will use Gaussian weights, 
%%
%\begin{equation}
%w_i(x) = \exp \left( - \frac{\|\Phi_{k-1}(x) - \Phi_{k-1} (x_i) \|^2}{\epsilon^2} \right), \text{where } \Phi_{k-1}(x) = \begin{bmatrix}
%\phi_1(x) \\
%\vdots \\
%\phi_{k-1}(x)
%\end{bmatrix} .
%\end{equation}
%%
%We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(x)$.
%
%Let $\hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
%
%We then define the normalized leave-one-out cross-validation error as
%\begin{equation} \label{eq:cv_error}
%r_{k} = \sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i)) \right)^2
%\end{equation}
%\begin{equation} \label{eq:cv_error}
%r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\Phi_{k-1}(x_i)) \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (x_i) \right) }}
%\end{equation}
%
Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
%
A large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
%
Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we define $r_1 = 1$.

@@

\begin{figure}[t] 
\centering
\includegraphics[width=\textwidth]{detect_change_eigenvalues}
\caption{The first two eigenvalues from diffusion maps as a function of $\lambda$. Each point is averaged over $10$ simulation experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation. 
The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
\label{fig:detect_change}
\end{figure}

@@

To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
%
Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
%
We plot  
\begin{equation}
 \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
\end{equation}
when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first. 
%
This is plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$.
